### 2023-03-10 abstract.tex
Make changes and add notes.

\begin{abstract}
\noindent
Sequence alignment is an essential method in bioinformatics and the basis of many analyses, including phylogenetic inference, ancestral sequence reconstruction, and gene annotation.
Sequence artifacts and errors made in alignment reconstruction can impact downstream analyses leading to erroneous conclusions in comparative and functional genomic studies.
For example, abiological frameshifts and early stop codons are common artifacts found in protein coding sequences annotated in reference genomes.
While this is eventually fixed in the reference genomes of model organisms, many genomes contain theses artifacts, and researchers often discard large amounts of data in comparative genomic studies to prevent artifacts from impacting results.
To address this need, we present COATi, a statistical, codon-aware pairwise aligner that supports complex insertion-deletion models and can handle artifacts present in genomic data.
COATi will allow users to reduce the amount of discarded data while generating more accurate sequence alignments.
\end{abstract}

>>>>

\begin{abstract}
\noindent
Sequence alignment is an essential method in bioinformatics and the basis of many analyses, including phylogenetic inference, ancestral sequence reconstruction, and gene annotation.
Sequence artifacts and errors made in alignment reconstruction can impact downstream analyses leading to erroneous conclusions in comparative and functional genomic studies.
For example, abiological frameshifts and early stop codons are common artifacts found in protein coding sequences that have been annotated in reference genomes.
While such errors are eventually fixed in the reference genomes of model organisms, many genomes used by researchers contain theses artifacts, and researchers often discard large amounts of data in comparative genomic studies to prevent artifacts from impacting results.
To address this need, we present COATi, a statistical, codon-aware pairwise aligner that supports complex insertion-deletion models and can handle artifacts present in genomic data.
COATi allows users to reduce the amount of discarded data while generating more accurate sequence alignments.
\end{abstract}

### 2023-03-10 alignpair_letter.tex
Make changes and add notes.

% title goes here:
\begin{flushleft}
{\Large\textbf{COATi: statistical pairwise alignment of protein coding sequences}}
\newline
% authors go here:
\\
Juan J. Garcia Mesa\textsuperscript{1,2},
Ziqi Zhu\textsuperscript{1,3},
Reed A. Cartwright\textsuperscript{1,3,*}
\\
\bigskip
1 Biodesign Institute, Arizona State University
\\
2 Ira A. Schools of Engineering, Arizona State University
\\
3 School of Life Sciences, Arizona State University
\\
\bigskip
* cartwright@asu.edu

>>>>

% title goes here:
\begin{flushleft}
{\Large\textbf{COATi: statistical pairwise alignment of protein coding sequences}}
\newline
% authors go here:
\\
Juan J. Garcia Mesa\textsuperscript{1,2},
Ziqi Zhu\textsuperscript{1,3},
Reed A. Cartwright\textsuperscript{1,3,*}
\\
\bigskip
1 The Biodesign Institute, Arizona State University, Tempe, Arizona, USA
\\
2 Ira A.\ Fulton Schools of Engineering, Arizona State University, Tempe, Arizona, USA
\\
3 School of Life Sciences, Arizona State University, Tempe, Arizona, USA
\\
\bigskip
* cartwright@asu.edu

### 2023-03-10 introduction.tex
Add missing citations
Make changes and add notes.

Sequence alignment is a fundamental task in bioinformatics and a cornerstone step in comparative and functional genomic analysis \citembe{sequence_alignment_rosenberg_2009}. While sophisticated advances have been made, the challenge of alignment inference has not been fully solved \citembe{art_morrison_2015}.
%Modern sequence analysis began with the heuristic homology algorithms of \textcite{Needleman1970} and \textcite{identification_smith_1981} and, while the methods developed since then have improved, alignment inference is not a solved problem \citep{art_morrison_2015}. 
%
The alignment of protein coding DNA sequences is one such challenge, and a common approach to this problem is to perform alignment inference in amino-acid space (e.g. \citetwo{bininda2005transalign}{abascal2010translatorx}).
While this approach is an improvement over DNA models, it discards information, underperforms compared to alignment at the codon level, and fails in the presence of artifacts such as frameshifts and early stop codons.
Although some aligners incorporate codon substitution models, they do not support frameshifts or lack a statistical model.
In addition, while modeling indels to appear within codons is rare, this is often the case \citembetwo{zhu2022profiling}{taylor2004occurrence}.
Considering gaps to only appear between codons can result in missing the optimal alignment and inflate estimates of sequence divergence (Fig. \ref{fig:aln}).

@@

Uncorrected errors in the alignment stage can lead to erroneous results in comparative and functional genomic studies \citembe{estimates_schneider_2009}.
Current methods are ill-equipped to handle common artifacts in genomic data, requiring costly curation practices that discard significant amounts of information.
To address this problem, we present COATi, short for COdon-aware Alignment Transducer, a pairwise statistical aligner that incorporates codon substitution models and is robust to artifacts present in modern genomic data.

>>>>

Sequence alignment is a fundamental task in bioinformatics and a cornerstone step in comparative and functional genomic analysis \citembe{sequence_alignment_rosenberg_2009}. While sophisticated advancements have been made, the challenge of alignment inference has not been fully solved \citembe{art_morrison_2015}.
%Modern sequence analysis began with the heuristic homology algorithms of \textcite{Needleman1970} and \textcite{identification_smith_1981} and, while the methods developed since then have improved, alignment inference is not a solved problem \citep{art_morrison_2015}. 
%
The alignment of protein coding DNA sequences is one such challenge, and a common approach to this problem is to perform alignment inference in amino-acid space (e.g. \citetwo{bininda2005transalign}{abascal2010translatorx}).
While this approach is an improvement over DNA models, it discards information, underperforms compared to alignment at the codon level, and fails in the presence of artifacts such as frameshifts and early stop codons.
Although some aligners incorporate codon substitution models, they do not support frameshifts or lack a statistical model.
In addition, existing aligners typically force gaps to occur between codons, whereas in natural sequences, only about 40\% of indels occur between codons \citembetwo{taylor2004occurrence}{zhu2022profiling}.
This mismatch between aligner assumptions and biology can produce sub-optimal alignments and inflated estimates of sequence divergence (Fig.\ \ref{fig:aln}).

@@

Genome quality impacts conclusions drawn from comparative genomic studies, and uncorrected errors in the alignment stage can lead to erroneous results in comparative and functional genomic studies \citembethree{estimates_schneider_2009}{effect_fletcher_2010}{hubisz2011error}.
Genomes for model organisms often get refined over many iterations and achieve high quality with meticulously curated protein coding sequences.
In contrast, genomes for non-model organisms might only receive partial curation and typically have lower quality sequences and annotations.
These genomes often lack the amount of sequencing data needed to fix artifacts, including missing exons, erroneous mutations, and indels \citembe{jackman2018tigmint}.
%
When comparative and functional genomics studies include data from non-model organisms, care must be taken to identify and manage such artifacts; however,
current alignment methods are ill-equipped to handle common artifacts in genomic data, requiring costly curation practices that discard significant amounts of information.
To address this problem, we present COATi, short for COdon-aware Alignment Transducer, a pairwise statistical aligner that incorporates codon substitution models and is robust to artifacts present in modern genomic data.

### 2023-03-10 materials_methods.tex
Make changes and add notes.

Statistical alignment is typically performed using pairwise hidden Markov models (pair-HMMs), which have the ability to rigorously model molecular sequence evolution \citembe{bradley2007transducers}.
Pair-HMMs are computational machines with two output tapes that contain a finite number of states typically labeled match, insert, and delete that emit symbols (nucleotides or amino acids) to one or both tapes.
Each tape represents a sequence and a path through a pair-HMM is a possible pairwise alignment.
Conceptually, these machines generate two sequences ($X$ and $Y$) from an unknown ancestor and can calculate the probability that two sequences are related, represented by $P(X, Y)$ \citembe{yoon_2009_hmm}.

A limitation of pair-HMMs is the ability to only model the evolution of two related sequences from an unknown ancestor.
Finite-state transducers (FSTs) have similar benefits to pair-HMMs with the additional feature to generate a descendant sequence given an ancestral one.
FSTs consume symbols from an input tape and emit symbols to an output tape.
Properly weighted, an FST can calculate the probability that a descendant sequence $Y$ evolved from an ancestor sequence $X$, represented by $P(Y | X)$.
Furthermore, well-established algorithms for combining FSTs in different ways allow the design of complex models by combining simpler FSTs \citembe{bradley2007transducers}.
A powerful and versatile algorithm for comparative sequence analysis is composition, which consists of sending the output of one FST into the input of a second FST.
The model implemented in COATi is designed by composing smaller FSTs, each representing a specific process.

Genome quality impacts conclusions drawn from comparative genomic studies \citembethree{estimates_schneider_2009}{effect_fletcher_2010}{hubisz2011error}.
Genomes for model organisms often get refined over many iterations and achieve high quality with meticulously curated protein coding sequences.
In contrast, genomes for non-model organisms might only receive partial curation and typically have lower quality sequences and annotations.
These genomes often lack the amount of sequencing data needed to fix artifacts, including missing exons, erroneous mutations, and indels \citembe{jackman2018tigmint}.
FSTs and their powerful methods provide a well-suited framework to statistically align a sequence from a non-model organism against a sequence from a model organism.

COATi implements the pairwise alignment of a potentially lower-quality sequence against a high-quality sequence as a path through the Evolution FST (Fig. \ref{fig:evolution-fst}), based on existing transducers (e.g. \cite{holmes2001evolutionary}).
Here, COATi treats the high-quality (reference) sequence as the ``ancestor'' and the potentially lower-quality sequence as the ``descendant''.
This FST is the result of composing a substitution FST that encodes a codon model (Fig. \ref{fig:evolution-fst}-a) and an indel FST that models insertions and deletions, including frameshifts (Fig. \ref{fig:evolution-fst}-b).
A key innovation of this FST with respect to others is the combination of a codon substitution model with a nucleotide-based geometric indel model that allows gaps to occur at any position.

Composing both sequences with the Evolution FST results in the transducer of all possible alignments.
Any path through this FST represents a pairwise alignment, while the shortest path corresponds to the best alignment.
All FST operations in COATi, including model development, composition, search for the shortest path, and other optimization algorithms, are performed using the C++ openFST library \citembe{allauzen2007openfst}.
However, the Evolution FST has a large state space to keep track of codon substitution rates when codons might be interspersed with indel events. This additional state space increase the computational complexity of the alignment algorithm.

@@

Codon substitution models are uncommon in sequence aligners, despite their extensive use in phylogenetics.
COATi implements the Muse and Gaut (\citeyear{muse_gaut_1994}) codon model (codon-triplet-mg) and the Empirical Codon Model \citembe{kosiol_ECM_2007} (codon-triplet-ecm).
It also lets the user provide a codon substitution matrix.
The default FST model (codon-triplet-mg) does not allow early stop codons on the ancestor sequence, although it supports mutations to (early) stop codons under the assumption that these are artifacts common in low-quality data.

@@

A marginal substitution model is calculated from a standard substitution model by calculating the marginal probabilities that each ancestral codon produces specific descendant nucleotides at each reading frame positions.
%
Specifically, let
$P_\text{cod}\left(Y_0 \cdot Y_1 \cdot Y_2 \middle| X_0 \cdot X_1 \cdot X_2\right)$ represent transition probabilities from a standard codon model, and
%
\[
P_\text{mar}\left(Y_p = y \middle| X_0 \cdot X_1 \cdot X_2\right) =
\sum_{Y_0 \cdot Y_1 \cdot Y_2} I(Y_p = y) P_\text{cod}\left(Y_0 \cdot Y_1 \cdot Y_2 \middle| X_0 \cdot X_1 \cdot X_2\right)
\]
%
represent the marginal transition probabilities, where
$p \in \{0, 1, 2\}$ is the position of the descendant nucleotide relative to the ancestral reading frame and $I$ is an indicator function defined
%
$I(y) := \{ 1 \text{ if } y = Y_p\text{, } 0 \text{ otherwise}\}$.
%
COATi contains marginal models for both Muse and Gaut or the Empirical Codon Model, resulting in the marginal models codon-marginal-mg (default model) and codon-marginal-ecm.
These models emphasize the position where the substitution in a codon occurs, help restrict the effects of low-quality data in the descendant sequence, and allow more than one substitution per codon.
In combination with the indel model, alignment using the marginal model is implemented using dynamic programming.

>>>>

Statistical alignment is typically performed using pairwise hidden Markov models (pair-HMMs), which have the ability to rigorously model molecular sequence evolution \citembe{bradley2007transducers}.
Pair-HMMs are computational machines with two output tapes. Pair-HMMs contain a finite number of states---typically labeled match, insert, and delete---that emit symbols (nucleotides or amino acids) to one or both tapes.
Each tape represents a sequence and a path through a pair-HMM is a possible pairwise alignment.
Conceptually, these machines generate two sequences ($\text{X}$ and $\text{Y}$) from an unknown ancestor and can calculate the probability that two sequences are related, represented by $P(\text{X}, \text{Y})$ \citembe{yoon_2009_hmm}.

A limitation of pair-HMMs is that they only model the evolution of two related sequences from an unknown ancestor.
Finite-state transducers (FSTs) have similar benefits to pair-HMMs with the additional feature that can can model the generation a descendant sequence given an ancestral one.
FSTs consume symbols from an input tape and emit symbols to an output tape.
Properly weighted, an FST can calculate the probability that a descendant sequence, $\text{Y}$, evolved from an ancestral sequence, $\text{X}$, represented by $P(\text{Y} | \text{X})$.
Furthermore, well-established algorithms for combining FSTs in different ways allow the design of complex models by combining simpler FSTs \citembe{bradley2007transducers}.
A powerful and versatile algorithm for comparative sequence analysis is composition, which consists of sending the output of one FST into the input of a second FST.
COATi uses composition to derive a statistical alignment model from the combination of smaller FSTs, each representing a specific process.

COATi implements the pairwise alignment of a potentially lower-quality sequence against a high-quality sequence as a path through the Evolution FST (Fig.\ \ref{fig:evolution-fst}) \citep[c.f.][]{holmes2001evolutionary}.
Here, COATi treats the high-quality (reference) sequence as the ``ancestor'' and the potentially lower-quality sequence as the ``descendant''.
This FST is the result of composing a substitution FST that encodes a codon model (Fig.\ \ref{fig:evolution-fst}-a) and an indel FST that models insertions and deletions, including frameshifts (Fig.\ \ref{fig:evolution-fst}-b).
A key innovation of this FST with respect to others is the combination of a codon substitution model with a nucleotide-based geometric indel model that allows gaps to occur at any position.

Composing both sequences with the Evolution FST results in the transducer of all possible alignments.
Any path through this FST represents a pairwise alignment, while the shortest path (by weight) corresponds to the best alignment.
All FST operations in COATi, including model development, composition, search for the shortest path, and other optimization algorithms, are performed using the C++ openFST library \citembe{allauzen2007openfst}.
However, the Evolution FST has a large state space to keep track of codon substitution rates when codons might be interspersed with indel events. This additional state space increases the computational complexity of the alignment algorithm.

@@

Codon substitution models are uncommon in sequence aligners, despite their extensive use in phylogenetics.
COATi implements the Muse and Gaut (\citeyear{muse_gaut_1994}) codon model (codon-triplet-mg) and the Empirical Codon Model \citembe{kosiol_ECM_2007} (codon-triplet-ecm).
It also lets the user provide a codon substitution matrix.
The default FST model (codon-triplet-mg) does not allow early stop codons in the ancestor sequence; although, it does supports mutations to (early) stop codons under the assumption that these are artifacts common in low-quality data.

@@

A marginal substitution model is calculated from a standard substitution model by calculating the marginal probabilities that each ancestral codon produces specific descendant nucleotides at each reading frame position.
%
Specifically, let
$P_\text{cod}\left(Y_0 \cdot Y_1 \cdot Y_2 \middle| X_0 \cdot X_1 \cdot X_2\right)$ represent transition probabilities from a standard codon model, and
%
\[
P_\text{mar}\left(Y_p = y \middle| X_0 \cdot X_1 \cdot X_2\right) =
\sum_{Y_0 \cdot Y_1 \cdot Y_2} I(Y_p = y) P_\text{cod}\left(Y_0 \cdot Y_1 \cdot Y_2 \middle| X_0 \cdot X_1 \cdot X_2\right)
\]
%
represent the marginal transition probabilities, where
$p \in \{0, 1, 2\}$ is the position of the descendant nucleotide relative to the ancestral reading frame and $I$ is an indicator function defined
%
$I(e) = \{ 1 \text{ if $e$ is true and } 0 \text{ otherwise}\}$.
%
COATi contains marginal models for both Muse and Gaut or the Empirical Codon Model, resulting in the marginal models codon-marginal-mg (default model) and codon-marginal-ecm.
These models emphasize the position in a codon where the substitution occurs, help restrict the effects of low-quality data in the descendant sequence, and allow more than one substitution per codon.
In combination with the indel model, alignment using the marginal model is implemented using dynamic programming.

### 2023-03-10 results_discussion.tex
Make changes and add notes.

\section*{Results and Discussion}
Using 16000 human genes and their gorilla homologous pairs from the ENSEMBL
database \citembe{ensembl_hubbard_2002}, we simulated a data set of pairwise
alignments with empirical gap patterns.
We used the data set to evaluate the accuracy of popular cutting edge aligners
Clustal$\Omega$ v1.2.4 \citembe{clustal_omega_sievers_2011},
MACSE v2.06 \citembe{ranwez_macse_2011}, MAFFT v7.407
\citembe{mafft_katoh_2002}, and PRANK v.170427 \citembe{prank_loytynoja_2014}
together with COATi.

After downloading, we removed 2232 sequences longer than 6000 nucleotides, identified 8369 sequence pairs that contained gaps identified by at least one aligner, and 5399 ungapped sequence pairs.
We then randomly introduced gap patterns extracted from all five methods into the ungapped sequence pairs to generate the benchmark alignments.
Alignment accuracy was measured using the distance metric $d_{seq}$
\citembe{metrics_blackburne_whelan_2011} between simulated and inferred
alignments.
In addition, accuracy of positive and negative selection was calculated
using the $F_1$ score by estimating $k_s$ and $k_a$ statistics
\citembe{ka_ks_li_1993}.

@@

>>>>

\section*{Results and Discussion}
Using 16000 human genes and their gorilla orthologs from the ENSEMBL
database \citembe{ensembl_hubbard_2002}, we simulated a data set of pairwise
alignments with empirical gap patterns.
We used the data set to evaluate the accuracy of popular aligners
Clustal$\Omega$ v1.2.4 \citembe{clustal_omega_sievers_2011},
MACSE v2.06 \citembe{ranwez_macse_2011}, MAFFT v7.407
\citembe{mafft_katoh_2002}, and PRANK v.170427 \citembe{prank_loytynoja_2014}
together with COATi. \TODO{which version of coati are we using. Should we show both? Or show both in a supplement?}

After downloading, we removed 2232 sequences longer than 6000 nucleotides, identified 8369 sequence pairs that contained gaps identified by at least one aligner, and 5399 ungapped sequence pairs.
We then randomly introduced gap patterns extracted from all five methods into the ungapped sequence pairs to generate the benchmark alignments.
Alignment accuracy was measured using the distance metric $d_{seq}$
\citembe{metrics_blackburne_whelan_2011} between simulated and inferred
alignments.
In addition, accuracy of positive and negative selection was calculated
using the $F_1$ score by estimating $k_s$ and $k_a$ statistics
\citembe{ka_ks_li_1993}. \TODO{briefly explain F1 score}

@@

\TODO{Add one short paragraph about future work. E.g. including a three-mer based indel model, and weighting indel phases differently.}
